%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Embedding Model Selection}
\label{ch:model-selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter covers model selection for the specific context of this repo:

\begin{itemize}
    \item Local inference centered on \texttt{llama.cpp} / \texttt{llama-server}
    \item The main compatibility target is an OpenAI-compatible local endpoint
    \item Open WebUI and similar clients are easiest to support when the embedding model
          is served as a dedicated local embeddings endpoint
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Status}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{warningbox}[Under Re-evaluation]
The original recommendation (Qwen3-Embedding-4B) was based on MTEB scores. After reviewing
RTEB research (see Chapter~\ref{ch:rteb}), the question of whether Qwen3-Embedding's rankings
reflect genuine out-of-distribution generalization or benchmark proximity remains open. The
final recommendation below reflects the current best assessment given available evidence.
\end{warningbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Shortlist}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Qwen3-Embedding-4B-GGUF}

\textbf{Key findings from the model card:}

\begin{itemize}
    \item 4B parameters, 32k context length
    \item Embedding dimension up to 2560, configurable output from 32 to 2560 dims
    \item 100+ language support
    \item Official GGUF quantizations including Q4\_K\_M
    \item Explicit \texttt{llama.cpp} usage documented on the model card:
\begin{Verbatim}
llama-embedding ... --pooling last
llama-server -m model.gguf --embedding --pooling last
\end{Verbatim}
    \item 69.45 on multilingual MTEB, 74.60 on MTEB English v2
\end{itemize}

\textbf{Practical footprint:}

\begin{itemize}
    \item Q4\_K\_M: \textasciitilde2.5 GB
    \item F16: \textasciitilde8.05 GB
\end{itemize}

\textbf{Fit assessment:} Best balance of quality, local footprint, and implementation
simplicity. Easiest path to adopt because it matches the existing llama.cpp serving model.

\subsection{Qwen3-Embedding-0.6B-GGUF}

\textbf{Key findings:}

\begin{itemize}
    \item 0.6B parameters, 32k context length, embedding dimension up to 1024
    \item 100+ language support, instruction awareness, configurable output dimension
    \item Q8\_0: \textasciitilde639 MB, F16: \textasciitilde1.2 GB
\end{itemize}

\textbf{Fit assessment:} Strongest low-footprint candidate. Best choice when memory or startup
speed matters more than absolute retrieval quality.

\subsection{Qwen3-Embedding-8B-GGUF}

\textbf{Key findings:}

\begin{itemize}
    \item 8B parameters, official GGUF release
    \item Q4\_K\_M: \textasciitilde4.68 GB
    \item Part of the same Qwen3 embedding family with multilingual, long-context, and
          code retrieval emphasis
\end{itemize}

\textbf{Fit assessment:} Strongest quality option in the reviewed Qwen family. Probably
overkill for a first local deployment unless retrieval quality is the top priority and the
extra memory cost is acceptable.

\subsection{BAAI/bge-m3}

\textbf{Key findings:}

\begin{itemize}
    \item Supports dense, sparse, and multi-vector (ColBERT) retrieval in one model
    \item 100+ languages, 8{,}192 token context
    \item The card explicitly recommends hybrid retrieval plus reranking for RAG
\end{itemize}

\textbf{Fit assessment:} Technically very capable and a strong retrieval baseline. Less
natural fit for this repo than Qwen3 GGUF because the Hugging Face card is centered on the
broader BGE ecosystem rather than a simple official GGUF + llama-server serving path.

This model is strongest when you are prepared to build a richer retrieval stack around dense
+ sparse + reranking, not when you want the cleanest single-model local embeddings server.

\subsection{mixedbread-ai/mxbai-embed-large-v1}

\textbf{Key findings:}

\begin{itemize}
    \item Strong English embedding model
    \item Retrieval queries should use the prompt:
          \textit{``Represent this sentence for searching relevant passages:''}
    \item Supports Matryoshka representation learning and quantization
    \item Strong March~2024 MTEB results; compares favorably with \texttt{text-embedding-3-large}
\end{itemize}

\textbf{Fit assessment:} A serious option for English-heavy retrieval, but adopting it
cleanly here would likely mean introducing a separate embeddings-serving stack (Infinity,
sentence-transformers) instead of reusing the local GGUF tooling pattern.

\subsection{nomic-embed-text-v1.5-GGUF}

\textbf{Key findings:}

\begin{itemize}
    \item Official GGUF release, explicitly llama.cpp-compatible
    \item Requires task instruction prefixes such as \texttt{search\_query:}
    \item Noted context-extension requirement for the full 8{,}192-token behavior
    \item Q4\_K\_M: \textasciitilde81 MiB
\end{itemize}

\textbf{Fit assessment:} Very attractive if tiny footprint is the main requirement. Weaker
choice than Qwen3 for a new deployment because it is older and its required task prefixes
add operational sharp edges.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary Table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{center}
\begin{tabular}{llll}
\toprule
\textbf{Model} & \textbf{Strength} & \textbf{Weakness} & \textbf{Fit} \\
\midrule
Qwen3-Embedding-4B & Quality + practicality & Heavier than tiny models & Best overall \\
Qwen3-Embedding-0.6B & Lightweight & Lower ceiling than 4B/8B & Best lightweight \\
Qwen3-Embedding-8B & Highest quality & More memory & Best max-quality \\
BAAI/bge-m3 & Dense+sparse versatility & Less direct fit & Good advanced \\
mxbai-embed-large-v1 & Strong English retrieval & Not GGUF-first & Separate stack \\
nomic-embed-text-v1.5 & Extremely small & Older, prefix-sensitive & Tiny fallback \\
\bottomrule
\end{tabular}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Final Recommendation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

If the goal is ``state of the art and compatible for local use'':

\begin{enumerate}
    \item Adopt \texttt{Qwen/Qwen3-Embedding-4B-GGUF}
    \item Serve it separately from the chat model
    \item Expose it as the local embeddings endpoint for RAG and Open WebUI
\end{enumerate}

Why this recommendation is defensible:

\begin{itemize}
    \item Qwen's Hugging Face cards present the strongest current benchmark story among
          reviewed candidates
    \item The family has an official GGUF distribution
    \item The card explicitly documents \texttt{llama.cpp} / \texttt{llama-server} usage
    \item The 4B size is large enough to be serious and small enough to be practical
    \item The Qwen3-Embedding family pairs with Qwen3-Reranker models for a complete
          two-stage retrieval stack
\end{itemize}

\begin{notebox}[Reranker Pairing]
Good retrieval systems often use both an embedding model for fast first-pass candidate
retrieval and a reranker for slower but more accurate second-pass ordering.

The Qwen3-Reranker family maps directly to the embedding family:
\begin{itemize}
    \item \texttt{Qwen/Qwen3-Reranker-0.6B} --- lightweight
    \item \texttt{Qwen/Qwen3-Reranker-4B} --- balanced
    \item \texttt{Qwen/Qwen3-Reranker-8B} --- maximum quality
\end{itemize}
\end{notebox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Recommended Next Steps}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{enumerate}
    \item Check whether Qwen3-Embedding RTEB scores have been published since this was
          written (the RTEB leaderboard is live and updated).
    \item Run a small domain-specific internal eval against your actual retrieval corpus.
          Even 50--100 query-document pairs with manual relevance labels will reveal gross
          distribution mismatch.
    \item If internal eval is not feasible, treat Qwen3-Embedding as the leading candidate
          but defer to RTEB-confirmed alternatives if they appear.
    \item If deploying only one model first, the embedding model is the right first step
          because it gives the core \texttt{/v1/embeddings} capability and is the prerequisite
          for scalable corpus search.
\end{enumerate}
