%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Retrieval Backends}
\label{ch:backends}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This chapter compares storage backends for medium-scale RAG and agent memory use cases.
Primary reference: analysis of \texttt{\texttildelow/d/memvid} (memvid-core v2.0.139, Rust)
conducted April~2026.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{memvid v2 --- Deep Analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{What It Is}

A single \texttt{.mv2} binary file containing everything: compressed payloads (zstd/lz4),
a WAL, a Tantivy BM25 index, an HNSW or flat vector index (bincode-serialized), a time index,
a CLIP visual index, an SPO entity-relationship graph, and typed ``MemoryCard'' facts.
No sidecar files. The entire corpus is one portable, optionally-encrypted artifact.

The name is a metaphor---there is no video codec. ``Frames'' are immutable append-only content
units inspired by video frame semantics. The v1 library (deprecated Python) used QR codes
embedded in MP4 files; v2 discards that entirely.

\subsection{Storage and Embedding Mechanics}

Embeddings are stored as raw \texttt{Vec<f32>} in the WAL at ingest time, materialized into
the vector index on \texttt{commit()}.

Index selection at build time:

\begin{itemize}
    \item \textbf{$<$ 1{,}000 vectors} $\to$ flat scan (bincode \texttt{Vec<VecDocument>}, linear L2 with SIMD)
    \item \textbf{$\geq$ 1{,}000 vectors} $\to$ HNSW (M=16, ef\_construction=200, hardcoded constants)
    \item \textbf{Optional PQ96} $\to$ 16$\times$ compression via product quantization, needs $\geq$ 100 vectors
\end{itemize}

Distance metric is \textbf{L2 (Euclidean)} throughout. Similarity score $= 1.0 - \text{distance}$.
Callers must normalize vectors externally for cosine equivalence.

Supported embedding sources:

\begin{itemize}
    \item Local ONNX (BGE-small 384d, BGE-base 768d, nomic-embed-text 768d, gte-large 1024d)
    \item OpenAI API (text-embedding-3-small 1536d, text-embedding-3-large 3072d, ada-002 1536d)
    \item CLIP visual (MobileCLIP-S2, 512d)
    \item Whisper audio (transcription $\to$ text $\to$ embed)
    \item Caller-provided \texttt{Vec<f32>} via \texttt{put\_with\_embedding()}
\end{itemize}

\subsection{Retrieval Patterns}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Method} & \textbf{What it does} \\
\midrule
\texttt{search()} / \texttt{search\_lex()} & BM25 via Tantivy with sketch pre-filter \\
\texttt{search\_vec()} & Pure ANN (flat or HNSW), returns distance scores \\
\texttt{search\_adaptive()} & Vector with score cutoff strategies (cliff, elbow) \\
\texttt{ask()} & Hybrid BM25 + vector with RRF ($k$=60), intent detection \\
\texttt{search\_clip()} & Visual similarity over CLIP embeddings \\
\texttt{graph\_search()} & Entity-relationship pattern matching over SPO MemoryCards \\
\texttt{timeline()} & Chronological retrieval with time-travel support \\
\bottomrule
\end{tabular}
\end{center}

\texttt{ask()} adjusts \texttt{top\_k} multipliers by intent: 1$\times$ (default),
2$\times$ (recency), 3$\times$ (aggregation), 5$\times$ (analytical).

\subsection{Memory and Agent Features}

\texttt{MemoryCard} types: facts, preferences, events, relationships. Populated via:

\begin{itemize}
    \item Explicit \texttt{put\_memory\_card()} API call
    \item Auto-extraction by enrichment workers (background, feature-gated)
    \item SPO triplet extractor that parses Subject-Predicate-Object from content
\end{itemize}

\texttt{Logic-Mesh} (entity graph) enables \texttt{graph\_search()} for natural language
entity queries (``who works at'', ``who likes'', ``lives in'').

\subsection{Performance Profile}

From their own benchmarks on local retrieval:

\begin{itemize}
    \item P50: \textbf{0.025ms}
    \item P99: \textbf{0.075ms}
\end{itemize}

\subsection{Hard Limitations}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Limitation} & \textbf{Impact} \\
\midrule
HNSW deletions are no-ops & Deleted frames remain in vector index \\
Cannot retrieve original embedding from HNSW/PQ & No export without full rebuild \\
PQ96 and HNSW params hardcoded & No recall/speed tuning \\
No pre-filtering during ANN & Metadata filters applied after retrieval only \\
Fully synchronous API & Requires \texttt{spawn\_blocking} in async Rust \\
Temporal TZ hardcoded to \texttt{America/Chicago} & Incorrect timestamps for non-US deployments \\
No distributed mode & Single-machine ceiling \\
\bottomrule
\end{tabular}
\end{center}

The HNSW deletion no-op is the most significant production concern. Any corpus needing
forgetting/eviction semantics must call \texttt{finalize\_indexes()} (full-scan rebuild) to
actually purge deleted embeddings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Backend Comparison Matrix}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{center}
\begin{tabular}{lllllll}
\toprule
& \textbf{memvid} & \textbf{Qdrant} & \textbf{LanceDB} & \textbf{pgvector} & \textbf{Chroma} & \textbf{sqlite-vec} \\
\midrule
Deployment & \texttt{.mv2} file & Docker & Embedded & PG extension & Server/embedded & SQLite ext \\
Self-hosted & Trivial & Easy & Trivial & Easy & Easy & Trivial \\
Hybrid search & BM25+vec+RRF & Sparse+dense & vec only & Manual FTS+vec & vec only & vec only \\
Pre-filter ANN & No & Yes & Yes & Approx & Post only & SQL WHERE \\
Deletion & HNSW: no-op & Yes & Yes (versioned) & Yes & Yes & Yes \\
Scale ceiling & 5--20M & 100M+ & 100M+ & 50M+ & $<$5M & $<$10M \\
Agent/memory & MemoryCards, SPO & None & None & None & None & None \\
\bottomrule
\end{tabular}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Key Backend Details}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Qdrant}

Self-hosted via Docker, Rust-native, disk-based HNSW with correct deletions. Inline payload
filtering during ANN search (exact pre-filter at small scale, sampling at large scale).
Sparse + dense hybrid search using FastEmbed for on-device BM25. Named collections map cleanly
to per-model or per-agent namespaces.

Quantization options are configurable at index creation time (scalar int8, PQ, binary).
REST and gRPC APIs. Strong operational tooling (Prometheus, dashboard).

\textbf{Best fit for:} medium-scale RAG with frequent updates, metadata-filtered ANN,
self-hosted.

\subsection{LanceDB}

Embedded (zero server), Rust-based, Apache Arrow + Lance columnar format. Supports S3 and
local disk. Versioned snapshots give correct delete/update semantics without a separate server
process. Python and Rust APIs both async.

Scale ceiling is much higher than memvid (100M+ vectors) with disk-based storage. Does not
include BM25 natively---Tantivy integration is an optional add-on.

\textbf{Best fit for:} large-scale RAG, columnar/tabular data, S3-backed or edge deployment,
zero ops.

\subsection{MongoDB Atlas}

Vector search requires Atlas cloud or the \texttt{mongot} sidecar process. \textbf{Not
available in community \texttt{mongod}.} This is the primary gate for local-first deployments.

When available: HNSW with configurable similarity metrics, Atlas Search (Lucene-based BM25,
facets, fuzzy), hybrid retrieval via \texttt{\$unionWith} + RRF scoring pipeline.

\textbf{Best fit for:} heterogeneous memory shapes, complex query logic, teams already using
Atlas, willing to accept cloud dependency or mongot overhead.

\subsection{pgvector}

PostgreSQL extension. Adds \texttt{vector} type, cosine operator, IVFFlat and HNSW indexes.
Standard SQL for filtering, joining, and ranking.

\textbf{Best fit for:} teams already running Postgres, structured metadata + vector hybrid
queries, medium scale ($<$50M vectors with tuning).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Decision Guide}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Scenario} & \textbf{Pick} \\
\midrule
Agent long-term memory, time-travel, portable, air-gapped & \textbf{memvid} \\
Medium-scale RAG, updates/deletes, metadata-filtered ANN & \textbf{Qdrant} \\
Large-scale RAG, columnar, S3-backed, zero-server & \textbf{LanceDB} \\
Heterogeneous memory shapes, complex queries, Atlas acceptable & \textbf{MongoDB Atlas} \\
Existing Postgres infrastructure & \textbf{pgvector} \\
Prototype or small-scale Python RAG & \textbf{Chroma} \\
Ultra-lightweight single-file + SQL & \textbf{sqlite-vec} \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Stacking memvid with Another Backend}

memvid's agent memory features (MemoryCards, SPO graph, timeline, time-travel) are unique and
not easily replicated. Its RAG limitations (HNSW deletion no-op, no pre-filtering) are real.

A viable hybrid: use \textbf{memvid for agent memory} (structured facts, conversation
history, time-travel replay) and \textbf{Qdrant for document RAG} (large corpus, filtered
ANN, correct deletes). Route queries through clawrouter with memory context injected from
memvid alongside document context from Qdrant.
